---
title: "W209 Final Project Data Munging - Wide to Long"
author: "Nathaniel Velarde"
date: "November 9, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
# Removing all of the variables in the current work space
rm(list = ls())

# setting global chunk options
library(knitr)

# *tidy up* the code when they are printed out
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=FALSE)
# removes the ## before output lines
opts_chunk$set(comment=NA)
# changing max.print option - default = 1000
options(max.print = 10000)
```

```{r libraries, echo=FALSE, include=FALSE}
library(tidyr)
library(dplyr) # includes dyplr, ggplot2, tidyr, tibble
library(data.table)
```


# Loading Data File - `EdStatsData.csv`

```{r}
# loading in data from csv - large file, takes +1 minute to load
df <- read.csv("EdStatsData.csv")
```

```{r}
# examining loaded data
dim(df)
cat("\n")
head(df)
```

```{r}
# getting column names
colnames(df)
```

# Eliminating Columns With All NAs

```{r}
# basic code to check if column has all na's
all(is.na(df$X))
```

```{r}
all_na_col <- sapply(df, function(x) all(is.na(x)))
```

**Used `sapply` versus `apply` and `lapply` because you can retrieve the column name and then use it for the results `data.frame`**

```{r}
all_na_col <- data.frame(all_na_col)
```

```{r}
# only one column has all NAs
head(all_na_col)
tail(all_na_col)
```

```{r}
dim(df)
```

```{r}
# going to drop the column with all NAs - X
df <- subset(df, select = -X)
dim(df)
cat("\n")
head(df)
```

# Removing Rows Where All Year Columns Have NAs

```{r}
colnames(df)
```

```{r}
# colnames gives you a character vector
# for this part of the munging, only interested in the years columns
colnames(df)[5:69]
```

```{r}
# proof of concept that can slice an r dataframe like this
head(df)[colnames(df)[5:69]]
```

```{r}
length(colnames(df)[5:69])
```

## Going to create a new dataframe consisting just of the year columns

```{r}
# colnmaes is a character vector
years_subframe <- df[colnames(df)[5:69]]

dim(years_subframe) # should have the same number of rows as df, but 4 less columns
dim(df)

cat("\n")

head(years_subframe)
```

## Now going to remove all rows in the years subframe that have all NAs

```{r}
years_row_na <- apply(years_subframe, 1, function(x) all(is.na(x)) )

length(years_row_na) # should have same number of rows as df
cat("\n")
head(years_row_na)
```

```{r}
# number of rows that have at least one non-NA value -- the number of rows in the new "cleaned" dataframe to
# be constructed below
sum(years_row_na)
```

```{r}
# number of rows and column in dataframe before NA row removal
dim(df)
cat("\n")
dim(df)[1] - sum(years_row_na) # number of all NA rows to be removed
```

## Use resulting vector to mask main dataframe

```{r}
# masking dataframe
df <- df[years_row_na == FALSE, ]

# confirming results
dim(df)
cat("\n")
head(df)
```

# Now can move to convert dataframe from wide to long format

Our ideal format is to create a new column `Year` which will designate the Year for each row and each indicator of interest will have its own column with values corresponding to a `country-year`

```{r}
wide <- df
```

## Going to use `dyplr::gather()` to convert to individual year columns to a single `Year` column and a `Value` column

- the `Value` column will then be expanded later such that each indicator will have its own column

```{r}
long <- wide %>% gather("Year", "Value", X1970:X2100)

dim(long)
cat("\n")
head(long)
```

**Removing X from Year**

```{r}
# removing the 'X' from Year aand converting datatype from string to numeric (which becomes useful below)
long$Year = as.numeric(gsub("X", "", long$Year))
head(long)
```

**At this point you could export this to a .csv file, but the file would be huge with 23.2 MM rows.  Limiting the data to only useful indicators and removing the forecast years -- feel free to leave the forecast years in if you find a forecast indicator with values for many countries**

## Filtering out Forecast Years

```{r}
# taking out years 2020 and greater
long <- long %>% filter(Year < 2020)
dim(long)
```

## Creating a `filter-vector` of the indicators of interest

**Easier to use the `Indicator.Code` because you have to create the `filter.vector` manually**

**These are the ones I chose, you can select your own -- easiest to cut and paste the `Indicator.Code` from Excel**

```{r}
filter_vector <- c("SE.PRE.ENRL", "SE.PRE.ENRL.FE",
                "SE.PRM.ENRL", "SE.PRM.ENRL.FE",
                "SE.SEC.ENRL", "SE.SEC.ENRL.FE",
                "SE.TER.ENRL", "SE.TER.ENRL.FE")
```


```{r}
# of Indicators Chosen
length(filter_vector)
```

## Filtering `long` dataframe by `filter.vector`
```{r}
head(long)
```
```{r}
#set regions that want to keep
region = c('East Asia & Pacific', 'Europe & Central Asia', 'Latin America & Caribbean', 'Middle East & North Africa', 'North America', 'Sub-Saharan Africa')

income = c('High income', 'Upper middle income', 'Middle income','Lower middle income','Low income')

```

```{r}
#filtered_long = long[long$Indicator.Code == filter_vector]
filtered_1 <- subset(long, Indicator.Code == "SE.PRE.ENRL", select = c("Country.Name","Country.Code","Indicator.Name","Year","Value"))
filtered_1$Education = "Pre-primary"
filtered_1$Gender = "Both"
head(filtered_1)
dim(filtered_1)
```

```{r}
library(tidyr)

filtered_1_wide = spread(filtered_1, Indicator.Name, Value)
setnames(filtered_1_wide, old = c('Country.Name','Country.Code', 'Enrolment in pre-primary education, both sexes (number)'), new = c('Region.Name','Region.Code', 'Number.of.Enrolment'))
filtered_1_wide <- filtered_1_wide %>% filter(Region.Name %in% region)
head(filtered_1_wide)
```

```{r}
#do all the other 7 indicators
filtered_2 <- subset(long, Indicator.Code == "SE.PRE.ENRL.FE", select = c("Country.Name","Country.Code","Indicator.Name","Year","Value"))
filtered_2$Education = "Pre-primary"
filtered_2$Gender = "Female"
filtered_2_wide = spread(filtered_2, Indicator.Name, Value)
setnames(filtered_2_wide, old = c('Country.Name','Country.Code', 'Enrolment in pre-primary education, female (number)'), new = c('Region.Name','Region.Code', 'Number.of.Enrolment'))
filtered_2_wide <- filtered_2_wide %>% filter(Region.Name %in% region)

filtered_3 <- subset(long, Indicator.Code == "SE.PRM.ENRL", select = c("Country.Name","Country.Code","Indicator.Name","Year","Value"))
filtered_3$Education = "Primary"
filtered_3$Gender = "Both"
filtered_3_wide = spread(filtered_3, Indicator.Name, Value)
setnames(filtered_3_wide, old = c('Country.Name','Country.Code', 'Enrolment in primary education, both sexes (number)'), new = c('Region.Name','Region.Code', 'Number.of.Enrolment'))
filtered_3_wide <- filtered_3_wide %>% filter(Region.Name %in% region)

filtered_4 <- subset(long, Indicator.Code == "SE.PRM.ENRL.FE", select = c("Country.Name","Country.Code","Indicator.Name","Year","Value"))
filtered_4$Education = "Primary"
filtered_4$Gender = "Female"
filtered_4_wide = spread(filtered_4, Indicator.Name, Value)
setnames(filtered_4_wide, old = c('Country.Name','Country.Code', 'Enrolment in primary education, female (number)'), new = c('Region.Name','Region.Code', 'Number.of.Enrolment'))
filtered_4_wide <- filtered_4_wide %>% filter(Region.Name %in% region)

filtered_5 <- subset(long, Indicator.Code == "SE.SEC.ENRL", select = c("Country.Name","Country.Code","Indicator.Name","Year","Value"))
filtered_5$Education = "Secondary"
filtered_5$Gender = "Both"
filtered_5_wide = spread(filtered_5, Indicator.Name, Value)
setnames(filtered_5_wide, old = c('Country.Name','Country.Code', 'Enrolment in secondary education, both sexes (number)'), new = c('Region.Name','Region.Code', 'Number.of.Enrolment'))
filtered_5_wide <- filtered_5_wide %>% filter(Region.Name %in% region)

filtered_6 <- subset(long, Indicator.Code == "SE.SEC.ENRL.FE", select = c("Country.Name","Country.Code","Indicator.Name","Year","Value"))
filtered_6$Education = "Secondary"
filtered_6$Gender = "Female"
filtered_6_wide = spread(filtered_6, Indicator.Name, Value)
setnames(filtered_6_wide, old = c('Country.Name','Country.Code', 'Enrolment in secondary education, female (number)'), new = c('Region.Name','Region.Code', 'Number.of.Enrolment'))
filtered_6_wide <- filtered_6_wide %>% filter(Region.Name %in% region)

filtered_7 <- subset(long, Indicator.Code == "SE.TER.ENRL", select = c("Country.Name","Country.Code","Indicator.Name","Year","Value"))
filtered_7$Education = "Tertiary"
filtered_7$Gender = "Both"
filtered_7_wide = spread(filtered_7, Indicator.Name, Value)
setnames(filtered_7_wide, old = c('Country.Name','Country.Code', 'Enrolment in tertiary education, all programmes, both sexes (number)'), new = c('Region.Name','Region.Code', 'Number.of.Enrolment'))
filtered_7_wide <- filtered_7_wide %>% filter(Region.Name %in% region)

filtered_8 <- subset(long, Indicator.Code == "SE.TER.ENRL.FE", select = c("Country.Name","Country.Code","Indicator.Name","Year","Value"))
filtered_8$Education = "Tertiary"
filtered_8$Gender = "Female"
filtered_8_wide = spread(filtered_8, Indicator.Name, Value)
setnames(filtered_8_wide, old = c('Country.Name','Country.Code', 'Enrolment in tertiary education, all programmes, female (number)'), new = c('Region.Name','Region.Code', 'Number.of.Enrolment'))
filtered_8_wide <- filtered_8_wide %>% filter(Region.Name %in% region)
```


```{r}
filtered_both = rbind(filtered_1_wide,filtered_3_wide,filtered_5_wide,filtered_7_wide)
filtered_female = rbind(filtered_2_wide,filtered_4_wide,filtered_6_wide,filtered_8_wide)
filtered_male =  merge(filtered_both,filtered_female,by=c("Region.Name","Region.Code","Year","Education"))
filtered_male$Gender = "Male"
filtered_male$Number.of.Enrolment = filtered_male$Number.of.Enrolment.x - filtered_male$Number.of.Enrolment.y
filtered_male$Gender.x = NULL
filtered_male$Number.of.Enrolment.x = NULL
filtered_male$Gender.y = NULL
filtered_male$Number.of.Enrolment.y = NULL
```

```{r}
filtered_all = rbind(filtered_female,filtered_male)
filtered_all_education = filtered_all
filtered_all_education$Vizside = "education"
filtered_all_gender = filtered_all
filtered_all_gender$Vizside = "gender"
filtered_all_region = filtered_all
filtered_all_region$Vizside = "region"

filtered_all_all = rbind(filtered_all_education,filtered_all_gender,filtered_all_region)

```

# Writing to CSV

```{r}
write.csv(filtered_all_all, file = "sankey_data.csv", row.names=FALSE)
```








