---
title: "W209 Final Project Data Munging - Wide to Long - Version 4"
author: "Nathaniel Velarde"
date: "December 2, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
# Removing all of the variables in the current work space
rm(list = ls())

# setting global chunk options
library(knitr)

# *tidy up* the code when they are printed out
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=FALSE)
# removes the ## before output lines
opts_chunk$set(comment=NA)
# changing max.print option - default = 1000
options(max.print = 10000)
```

```{r libraries, echo=FALSE, include=FALSE}
library(tidyr)
library(plyr) # need to load plyr before dplyr
library(dplyr) # includes dyplr, ggplot2, tidyr, tibble
library(data.table)
```


# Loading Data File - `EdStatsData.csv`

```{r}
# loading in data from csv - large file, takes +1 minute to load
df <- read.csv("EdStatsData.csv")
```

```{r}
# examining loaded data
dim(df)
cat("\n")
head(df)
```

```{r}
# getting column names
colnames(df)
```

# Eliminating Columns With All NAs

```{r}
# basic code to check if column has all na's
all(is.na(df$X))
```

```{r}
all_na_col <- sapply(df, function(x) all(is.na(x)))
```

**Used `sapply` versus `apply` and `lapply` because you can retrieve the column name and then use it for the results `data.frame`**

```{r}
all_na_col <- data.frame(all_na_col)
```

```{r}
# only one column has all NAs
head(all_na_col)
tail(all_na_col)
```

```{r}
dim(df)
```

```{r}
# going to drop the column with all NAs - X
df <- subset(df, select = -X)
dim(df)
cat("\n")
head(df)
```

# Removing Rows Where All Year Columns Have NAs

```{r}
colnames(df)
```

```{r}
# colnames gives you a character vector
# for this part of the munging, only interested in the years columns
colnames(df)[5:69]
```

```{r}
# proof of concept that can slice an r dataframe like this
head(df)[colnames(df)[5:69]]
```

```{r}
length(colnames(df)[5:69])
```

## Going to create a new dataframe consisting just of the year columns

```{r}
# colnames is a character vector
years_subframe <- df[colnames(df)[5:69]]

dim(years_subframe) # should have the same number of rows as df, but 4 less columns
dim(df)

cat("\n")

head(years_subframe)
```

## Now going to remove all rows in the years subframe that have all NAs

```{r}
years_row_na <- apply(years_subframe, 1, function(x) all(is.na(x)) )

length(years_row_na) # should have same number of rows as df
cat("\n")
head(years_row_na)
```

```{r}
# number of rows that have at least one non-NA value -- the number of rows in the new "cleaned" dataframe to
# be constructed below
sum(years_row_na)
```

```{r}
# number of rows and column in dataframe before NA row removal
dim(df)
cat("\n")
dim(df)[1] - sum(years_row_na) # number of all NA rows to be removed
```

## Use resulting vector to mask main dataframe

```{r}
# masking dataframe
df <- df[years_row_na == FALSE, ]

# confirming results
dim(df)
cat("\n")
head(df)
```

# Now can move to convert dataframe from wide to long format

Our ideal format is to create a new column `Year` which will designate the Year for each row and each indicator of interest will have its own column with values corresponding to a `country-year`

```{r}
wide <- df
```

## Going to use `dyplr::gather()` to convert to individual year columns to a single `Year` column and a `Value` column

- the `Value` column will then be expanded later such that each indicator will have its own column

```{r}
long <- wide %>% gather("Year", "Value", X1970:X2100)

dim(long)
cat("\n")
head(long)
```

**Removing X from Year**

```{r}
# removing the 'X' from Year aand converting datatype from string to numeric (which becomes useful below)
long$Year = as.numeric(gsub("X", "", long$Year))
head(long)
```

**At this point you could export this to a .csv file, but the file would be huge with 23.2 MM rows.  Limiting the data to only useful indicators and removing the forecast years -- feel free to leave the forecast years in if you find a forecast indicator with values for many countries**

## Filtering out Forecast Years

```{r}
# taking out years 2020 and greater
long <- long %>% filter(Year < 2020)
dim(long)
```

## Creating a `filter-vector` of the indicators of interest

**Easier to use the `Indicator.Code` because you have to create the `filter.vector` manually**

**These are the ones I chose, you can select your own -- easiest to cut and paste the `Indicator.Code` from Excel**

```{r}
filter_vector <- c("SE.PRM.ENRR", # Gross Enrolment Rate, Primary
                   "SE.PRM.ENRR.FE", # Gross Enrolment Rate, Primary, Female
                   "SE.PRM.ENRR.MA", # Gross Enrolment Rate, Primary, Male
                   "SE.SEC.ENRR", # Gross Enrolment Rate, Secondary
                   "SE.SEC.ENRR.FE", # Gross Enrolment Rate, Secondary, Female
                   "SE.SEC.ENRR.MA", # Gross Enrolment Rate, Secondary, Male
                   "SE.TER.ENRR", # Gross Enrolment Rate, Tertiary
                   "SE.TER.ENRR.FE", # Gross Enrolment Rate, Teriary, Female
                   "SE.TER.ENRR.MA", # Gross Enrolment Rate, Teriary, Male
                   "SE.ENR.PRIM.FM.ZS", # GPI, Gross Enrollment Ratio, Primary
                   "SE.ENR.SECO.FM.ZS", # GPI, Gross Enrollment Ratio, Secondary
                   "SE.ENR.TERT.FM.ZS", # GPI, Gross Enrollment Ratio, Teriary
                   "SP.POP.GROW", # Population Growth
                   "SP.POP.TOTL", # Total Population -- MANUAL CALC OF SP.POP.GROW
                   "NY.GDP.PCAP.KD", # per capita GDP Constant 2010 Dollars
                   "NY.GDP.PCAP.PP.KD" # per capita GDP, PPP, Constant 2011 Int'l Dollars
                   )
```


```{r}
# of Indicators Chosen
length(filter_vector)
```

## Filtering `long` dataframe by `filter.vector`

```{r}
filtered_long <- long %>% filter(Indicator.Code %in% filter_vector)

dim(filtered_long)
cat("\n")
head(filtered_long)
```

**At this point, you could export the filtered dataframe to .csv -- this format may be compatible with Tableau, but you may have to use the pivot function or some other manipulation to get it to work in Tableau**

**To export to .csv**

```{r}
# write.csv(df_name_to_export, file = "YOUR_FILENAME.csv", row.names=FALSE)
```



===========================================================

# Retaining a Single Column of Values -- `Indicator.Value` -- and `Indicator.Code` that may facillitate some Tableau Functions

```{r}
filtered_long$Indicator.Value <- filtered_long$Value
```

```{r}
filtered_long$Indicator.CodeSpread <- filtered_long$Indicator.Code
```

```{r}
head(filtered_long)
```

==================================================

# Now converting back to (Semi) Wide Format with each Indicator.Code Having its Own Column of Values

```{r}
# expect new df to have num cols = orig cols - 2 (Indicator.Code and Value) + num new cols (length of filter_vector)
filtered_wide <- spread(filtered_long, Indicator.CodeSpread, Value)
```

```{r}
# confirming transformation resulted in correct dimensions
length(filter_vector)
cat("\n")
dim(filtered_long)
cat("\n")
dim(filtered_wide)
```

```{r}
head(filtered_wide)
```

## How to Test Data Availability in Terms of Number of Years for a Particlar Indicator.Code and Country

```{r}
# to see year availability need to have column of interest in select() and filter()
filtered_wide %>% 
  select(Country.Name, Year, NY.GDP.PCAP.PP.KD) %>% 
  filter(Country.Name == 'France', NY.GDP.PCAP.PP.KD)
```

## Creating GDP per Capita Annual Growth Rate

```{r}
dim(filtered_wide)
```

```{r}
# making a copy
final_export <- filtered_wide
```

```{r}
dim(filtered_wide)
dim(final_export)
```

```{r}
head(final_export, 2)
```

```{r}
# Proof of Concept
final_export %>% 
  group_by(Country.Name) %>% 
  arrange(Indicator.Code, Year) %>% 
  select(Country.Name, Year, NY.GDP.PCAP.PP.KD, Indicator.Code, Indicator.Value) %>%
  filter(Country.Name == 'France') %>%
  mutate(Pct.Growth = (  ( (Indicator.Value / lag(Indicator.Value)) - 1) * 100) ) %>%
  data.frame() %>%
  head(5)
```

```{r}
# this works!!!
test.frame <- final_export %>% 
  group_by(Country.Name) %>% 
  arrange(Country.Name, Indicator.Code, Year) %>% 
  mutate(Pct.Growth = (  ( (Indicator.Value / lag(Indicator.Value)) - 1) * 100) ) %>%
  data.frame()
```

```{r}
# testing
test.frame %>% 
  select(Country.Name, Year, NY.GDP.PCAP.PP.KD, Indicator.Code, Indicator.Value, Pct.Growth) %>% 
  filter(Country.Name == 'France') %>%
  head()
```

```{r}
# testing population growth
test.frame %>% 
  select(Country.Name, Year, Indicator.Code, SP.POP.GROW, SP.POP.TOTL, Indicator.Value, Pct.Growth) %>% 
  filter(Country.Name == 'France', Indicator.Code == 'SP.POP.TOTL') %>%
  head(10)
```

```{r}
# testing population growth -- looking at DATASET SP.POP.GROW and compare to above Pct.Growth
test.frame %>% 
  select(Country.Name, Year, Indicator.Code, SP.POP.GROW, SP.POP.TOTL, Indicator.Value, Pct.Growth) %>% 
  filter(Country.Name == 'France', Indicator.Code == 'SP.POP.GROW') %>%
  head(10)
```

```{r}
# For 1971
(52480421 / 52035095 - 1) * 100
```

```{r}
test.frame %>% 
  filter(Country.Name == 'Canada', Indicator.Code == 'SE.TER.ENRR.FE') %>%
  select(Country.Name, Year, Indicator.Code, Indicator.Value) %>%
  head(10)
```

```{r}
# EXPORT VERSION
rmunged_data <- final_export %>% 
  group_by(Country.Name) %>% 
  arrange(Country.Name, Indicator.Code, Year) %>% 
  mutate(Pct.Growth = (  ( (Indicator.Value / lag(Indicator.Value)) - 1) * 100) ) %>%
  data.frame()
```

```{r}
dim(test.frame)
dim(rmunged_data)
```

```{r}
unique(rmunged_data$Indicator.Name)
```

```{r}
unique(rmunged_data$Indicator.Code)
```

```{r}
orig.code.names <- as.vector(unique(rmunged_data$Indicator.Code))
```

```{r}
orig.code.names
```

```{r}
head(rmunged_data, 2)
```

```{r}
old.names <- c("NY.GDP.PCAP.KD",
               "NY.GDP.PCAP.PP.KD",
               "SE.ENR.PRIM.FM.ZS",
               "SE.ENR.SECO.FM.ZS",
               "SE.ENR.TERT.FM.ZS",
               "SE.PRM.ENRR",
               "SE.PRM.ENRR.FE",
               "SE.PRM.ENRR.MA",
               "SE.SEC.ENRR",
               "SE.SEC.ENRR.FE",
               "SE.SEC.ENRR.MA",
               "SE.TER.ENRR",
               "SE.TER.ENRR.FE",
               "SE.TER.ENRR.MA",
               "SP.POP.GROW",
               "SP.POP.TOTL"
               )
```

```{r}
# Creating Human Readable Names
new.names <- c("GDP per Capita Constant $",
               "GDP per Capita PPP",
               "GPI-Primary",
               "GPI-Secondary",
               "GPI-Teriary",
               "Primary Enrollment-All",
               "Primary Enrollment-Female",
               "Primary Enrollment-Male",
               "Secondary Enrollment-All",
               "Secondary Enrollment-Female",
               "Secondary Enrollment-Male",
               "Teriary Enrollment-All",
               "Teriary Enrollment-Female",
               "Teriary Enrollment-Male",
               "Annual Pop Growth %",
               "Total Population"
               )
```

```{r}
length(orig.code.names)
length(old.names)
length(new.names)
```

# Creating Human Readable Indicator

```{r}
head(rmunged_data, 2)
```

```{r}
rmunged_data$Indicator <- mapvalues(rmunged_data$Indicator.Code,
                                    from = old.names,
                                    to = new.names)
```

```{r}
head(rmunged_data, 2)
```

```{r}
# confirming
rmunged_data %>% 
  select(Indicator, Indicator.Code, Indicator.Value) %>%
  filter(Indicator.Code == "SP.POP.TOTL") %>%
  head(20)
```

```{r}
filter_vector
```

```{r}
test.rmunge <- rmunged_data
```

```{r}
head(test.rmunge, 2)
```

```{r}
# Creating Human Readable Names - Educational Metrics
ed.names <- c("NA",
              "NA",
              "GPI-Primary",
              "GPI-Secondary",
              "GPI-Teriary",
              "Primary Enrollment-All",
              "Primary Enrollment-Female",
              "Primary Enrollment-Male",
              "Secondary Enrollment-All",
              "Secondary Enrollment-Female",
              "Secondary Enrollment-Male",
              "Teriary Enrollment-All",
              "Teriary Enrollment-Female",
              "Teriary Enrollment-Male",
              "NA",
              "NA"
               )
```

```{r}
# Creating Human Readable Names - Macro Indicators
macro.names <- c("GDP per Capita Constant $",
                 "GDP per Capita PPP",
                 "NA",
                 "NA",
                 "NA",
                 "NA",
                 "NA",
                 "NA",
                 "NA",
                 "NA",
                 "NA",
                 "NA",
                 "NA",
                 "NA",
                 "Annual Pop Growth %",
                 "Total Population"
                 )
```

```{r}
length(new.names)
length(ed.names)
length(macro.names)
```

# Creating Education.Indicators

```{r}
test.rmunge$Education.Indicators <- mapvalues(test.rmunge$Indicator,
                                    from = new.names,
                                    to = ed.names)
```

```{r}
head(test.rmunge, 2)
```


```{r}
# confirming
test.rmunge %>% 
  select(Country.Name, Indicator, Indicator.Code, Indicator.Value, Education.Indicators) %>%
  filter(Country.Name == 'France', Indicator.Code == "SP.POP.TOTL") %>%
  head(20)
```

# Creating Macro.Indicators

```{r}
test.rmunge$Macro.Indicators <- mapvalues(test.rmunge$Indicator,
                                          from = new.names,
                                          to = macro.names)
```

```{r}
head(test.rmunge, 2)
```

```{r}
# confirming
test.rmunge %>% 
  select(Country.Name, Indicator, Indicator.Code, Indicator.Value, Macro.Indicators) %>%
  filter(Country.Name == 'France', Indicator.Code == "SP.POP.TOTL") %>%
  head(20)
```

```{r}
# should be 2 new columns
dim(rmunged_data)
dim(test.rmunge)
```

```{r}
rmunged_data2 <- test.rmunge
```

```{r}
dim(rmunged_data2)
```

```{r}
head(rmunged_data2, 2)
```

```{r}
d <- rmunged_data2
```

```{r}
names(d)
```

## Creating Human Readable Column Names

```{r}
head(d, 2)
```

```{r}
names(d)[names(d) == "NY.GDP.PCAP.KD"] <- "GDP.Capita.Constant"
```

```{r}
head(d, 2)
```

```{r}
names(d)[names(d) == "NY.GDP.PCAP.PP.KD"] <- "GDP.Capita.PPP"
names(d)[names(d) == "SE.ENR.PRIM.FM.ZS"] <- "GPI.Primary"
names(d)[names(d) == "SE.ENR.SECO.FM.ZS"] <- "GPI.Secondary"
names(d)[names(d) == "SE.ENR.TERT.FM.ZS"] <- "GPI.Tertiary"
names(d)[names(d) == "SE.PRM.ENRR"] <- "Primary.Enrollment.All"
names(d)[names(d) == "SE.PRM.ENRR.FE"] <- "Primary.Enrollment.Female"
names(d)[names(d) == "SE.PRM.ENRR.MA"] <- "Primary.Enrollment.Male"
names(d)[names(d) == "SE.SEC.ENRR"] <- "Secondary.Enrollment.All"
names(d)[names(d) == "SE.SEC.ENRR.FE"] <- "Secondary.Enrollment.Female"
names(d)[names(d) == "SE.SEC.ENRR.MA"] <- "Secondary.Enrollment.Male"
names(d)[names(d) == "SE.TER.ENRR"] <- "Tertiary.Enrollment.All"
names(d)[names(d) == "SE.TER.ENRR.FE"] <- "Tertiary.Enrollment.Female"
names(d)[names(d) == "SE.TER.ENRR.MA"] <- "Tertiary.Enrollment.Male"
names(d)[names(d) == "SP.POP.GROW"] <- "Population.Growth"
names(d)[names(d) == "SP.POP.TOTL"] <- "Total.Population"
```

```{r}
head(d, 2)
```

# Will create composite columns in Tableau


# Writing to CSV - Simple (No Additional Filtering Columns)

```{r}
head(d)
```

```{r}
# writing to csv
# write.csv(d, file = "R_Munged_Final_A_Sunday.csv", row.names=FALSE)
```

======================================================================

# Adding Extra Columns

```{r}
d2 <- d
```

```{r}
dim(d)
dim(d2)
```

```{r}
# creating multiple YEARs for Tableau Filtering
d2$Year.X <- d2$Year
d2$Year.Y <- d2$Year
```

```{r}
# creating multiple Indicator.Value for Tableau Filtering
d2$Indicator.Value.X <- d2$Indicator.Value
d2$Indicator.Value.Y <- d2$Indicator.Value
```

```{r}
# creating multiple Indicator.Code for Tableau Filtering
d2$Indicator.Code.X <- d2$Indicator.Code
d2$Indicator.Code.Y <- d2$Indicator.Code
```

```{r}
# creating multiple Indicator (Human Readable) for Tableau Filtering
d2$Indicator.X <- d2$Indicator
d2$Indicator.Y <- d2$Indicator
```

```{r}
dim(d)
dim(d2)
```

```{r}
head(d2)
```

```{r}
d2 %>% 
  filter(Country.Name == 'Finland', Year == 2010) %>% 
  select(Country.Name, Year, Indicator.Code, Indicator.Value)
```

# Writing to .csv -- Complex -- Additional Filtering Columns

```{r}
# writing to csv
# write.csv(d2, file = "R_Munged_Final_B_Sunday.csv", row.names=FALSE)
```

=========================================================================================

=========================================================

# Verifying Composite Calculations

## Creating Test Dataframe

```{r}
test.d2 <- d2 %>% filter(Country.Name == 'France' | Country.Name == "Germany") %>% data.frame()
dim(test.d2)
```

```{r}
str(test.d2)
```

```{r}
head(test.d2, 2)
tail(test.d2, 2)
```

# Making Year a Factor So Can Group By

```{r}
test.d2$Year.Factor <- factor(test.d2$Year, ordered = TRUE)
```

```{r}
head(test.d2, 2)
```

# Verifying Composite GPI

```{r}
test.d2 %>% 
  filter(Country.Name == 'France' & Year.Factor == '1980') %>%
  select(Country.Name, Year.Factor, GPI.Primary, GPI.Secondary, GPI.Tertiary) %>% 
  data.frame %>% head(50)
```

```{r}
mean(c(0.98746, 1.11448, 0.8928))
```

## Matches

```{r}
mean(c(0.99825, 1.09390, 1.02507, 1.03342))
```

```{r}
mean(c(1.09390, 1.02507, 1.03342))
```

```{r}
mean(c(1.03601, 1.03846, 1.04267))
```

## Confirming Composite Enrollments

```{r}
test.d2 %>% 
  filter(Country.Name == 'France' & Year.Factor == '1980') %>%
  select(Country.Name, 
         Year.Factor, 
         Primary.Enrollment.Male, 
         Secondary.Enrollment.Male, 
         Tertiary.Enrollment.Male) %>% 
  data.frame %>% head(50)
```

```{r}
# All
mean(c(111.8651, 83.84619, 25.12804))
```

```{r}
# Female
mean(c(111.143, 88.50083, 23.67521))
```

```{r}
# Male
mean(c(112.5543, 79.41032, 26.51804))
```



```{r}
head(d2, 2)
```








```{r}
filter_vector
```



```{r}
test.d2 %>% 
  group_by(Country.Name, Year.Factor, GPI.Primary, GPI.Secondary, GPI.Tertiary) %>% 
  filter(Country.Name == 'France') %>%
  data.frame %>% head(50)
```
```






```{r}
dw <- read.table(header=T, text='
 sbj f1.avg f1.sd f2.avg f2.sd  blabla
   A   10    6     50     10      bA
   B   12    5     70     11      bB
   C   20    7     20     8       bC
   D   22    8     22     9       bD
 ')

dw
```


```{r}
dw %>% 
  gather(v, value, f1.avg:f2.sd) %>% 
  separate(v, c("var", "col")) %>% 
  arrange(sbj) %>% 
  spread(col, value)
```

```{r}

```




```{r}
stocks <- tibble(
  time = as.Date('2009-01-01') + 0:9,
  X = rnorm(10, 0, 1),
  Y = rnorm(10, 0, 2),
  Z = rnorm(10, 0, 4)
)
stocks %>% data.frame()
```

```{r}
gather(stocks, stock, price, -time) %>% data.frame()
#stocks %>% gather(stock, price, -time)
```

```{r}
stocks %>% gather(stock, price, -time)
```



