---
title: "W209 Final Project Data Munging - Wide to Long"
author: "Nathaniel Velarde"
date: "November 9, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
# Removing all of the variables in the current work space
rm(list = ls())

# setting global chunk options
library(knitr)

# *tidy up* the code when they are printed out
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=FALSE)
# removes the ## before output lines
opts_chunk$set(comment=NA)
# changing max.print option - default = 1000
options(max.print = 10000)
```

```{r libraries, echo=FALSE, include=FALSE}
library(tidyr)
library(dplyr) # includes dyplr, ggplot2, tidyr, tibble
library(data.table)
```


# Loading Data File - `EdStatsData.csv`

```{r}
# loading in data from csv - large file, takes +1 minute to load
df <- read.csv("EdStatsData.csv")
```

```{r}
# examining loaded data
dim(df)
cat("\n")
head(df)
```

```{r}
# getting column names
colnames(df)
```

# Eliminating Columns With All NAs

```{r}
# basic code to check if column has all na's
all(is.na(df$X))
```

```{r}
all_na_col <- sapply(df, function(x) all(is.na(x)))
```

**Used `sapply` versus `apply` and `lapply` because you can retrieve the column name and then use it for the results `data.frame`**

```{r}
all_na_col <- data.frame(all_na_col)
```

```{r}
# only one column has all NAs
head(all_na_col)
tail(all_na_col)
```

```{r}
dim(df)
```

```{r}
# going to drop the column with all NAs - X
df <- subset(df, select = -X)
dim(df)
cat("\n")
head(df)
```

# Removing Rows Where All Year Columns Have NAs

```{r}
colnames(df)
```

```{r}
# colnames gives you a character vector
# for this part of the munging, only interested in the years columns
colnames(df)[5:69]
```

```{r}
# proof of concept that can slice an r dataframe like this
head(df)[colnames(df)[5:69]]
```

```{r}
length(colnames(df)[5:69])
```

## Going to create a new dataframe consisting just of the year columns

```{r}
# colnmaes is a character vector
years_subframe <- df[colnames(df)[5:69]]

dim(years_subframe) # should have the same number of rows as df, but 4 less columns
dim(df)

cat("\n")

head(years_subframe)
```

## Now going to remove all rows in the years subframe that have all NAs

```{r}
years_row_na <- apply(years_subframe, 1, function(x) all(is.na(x)) )

length(years_row_na) # should have same number of rows as df
cat("\n")
head(years_row_na)
```

```{r}
# number of rows that have at least one non-NA value -- the number of rows in the new "cleaned" dataframe to
# be constructed below
sum(years_row_na)
```

```{r}
# number of rows and column in dataframe before NA row removal
dim(df)
cat("\n")
dim(df)[1] - sum(years_row_na) # number of all NA rows to be removed
```

## Use resulting vector to mask main dataframe

```{r}
# masking dataframe
df <- df[years_row_na == FALSE, ]

# confirming results
dim(df)
cat("\n")
head(test_df)
```

# Now can move to convert dataframe from wide to long format

Our ideal format is to create a new column `Year` which will designate the Year for each row and each indicator of interest will have its own column with values corresponding to a `country-year`

```{r}
wide <- df
```

## Going to use `dyplr::gather()` to convert to individual year columns to a single `Year` column and a `Value` column

- the `Value` column will then be expanded later such that each indicator will have its own column

```{r}
long <- wide %>% gather("Year", "Value", X1970:X2100)

dim(long)
cat("\n")
head(long)
```

**Removing X from Year**

```{r}
# removing the 'X' from Year aand converting datatype from string to numeric (which becomes useful below)
long$Year = as.numeric(gsub("X", "", long$Year))
head(long)
```

**At this point you could export this to a .csv file, but the file would be huge with 23.2 MM rows.  Limiting the data to only useful indicators and removing the forecast years -- feel free to leave the forecast years in if you find a forecast indicator with values for many countries**

## Filtering out Forecast Years

```{r}
# taking out years 2020 and greater
long <- long %>% filter(Year < 2020)
dim(long)
```

## Creating a `filter-vector` of the indicators of interest

**Easier to use the `Indicator.Code` because you have to create the `filter.vector` manually**

**These are the ones I chose, you can select your own -- easiest to cut and paste the `Indicator.Code` from Excel**

```{r}
filter_vector <- c("NY.GDP.PCAP.CD",
                   "SE.PRM.TENR",
                   "SE.PRM.TENR.FE",
                   "UIS.NERA.1.GPI",
                   "SE.PRM.DROP.ZS",
                   "SE.PRM.ENRR",
                   "SE.PRM.ENRR.FE",
                   "SE.ENR.PRIM.FM.ZS",
                   "SE.PRM.ENRR.MA",
                   "SE.SEC.ENRR",
                   "SE.SEC.ENRR.FE",
                   "SE.ENR.SECO.FM.ZS",
                   "SE.SEC.ENRR.MA",
                   "SE.TER.ENRR",
                   "SE.TER.ENRR.FE",
                   "SE.ENR.TERT.FM.ZS",
                   "SE.TER.ENRR.MA",
                   "SP.POP.GROW",
                   "SP.POP.0014.TO.ZS",
                   "SP.POP.1564.TO.ZS",
                   "SP.POP.TOTL.FE.ZS",
                   "SP.POP.TOTL.MA.ZS",
                   "BAR.NOED.15UP.FE.ZS",
                   "SE.XPD.TOTL.GB.ZS",
                   "SE.XPD.TOTL.GD.ZS",
                   "UIS.XUNIT.PPP.1.FSGOV",
                   "SL.TLF.ADVN.ZS",
                   "SL.TLF.ADVN.FE.ZS",
                   "SL.TLF.ADVN.MA.ZS",
                   "SL.TLF.BASC.ZS",
                   "SL.TLF.BASC.FE.ZS",
                   "SL.TLF.BASC.MA.ZS",
                   "SL.TLF.INTM.ZS",
                   "SL.TLF.INTM.FE.ZS",
                   "SL.TLF.INTM.MA.ZS",
                   "SL.TLF.TOTL.FE.ZS")
```


```{r}
# of Indicators Chosen
length(filter_vector)
```

## Filtering `long` dataframe by `filter.vector`

```{r}
filtered_long <- long %>% filter(Indicator.Code %in% filter_vector)

dim(filtered_long)
cat("\n")
head(filtered_long)
```

**At this point, you could export the filtered dataframe to .csv -- this format may be compatible with Tableau, but you may have to use the pivot function or some other manipulation to get it to work in Tableau**

**To export to .csv**

```{r}
# write.csv(df_name_to_export, file = "YOUR_FILENAME.csv", row.names=FALSE)
```

# Now converting back to (Semi) Wide Format with each Indicator.Code Having its Own Column of Values

```{r}
# expect new df to have num cols = orig cols - 2 (Indicator.Code and Value) + num new cols (length of filter_vector)
filtered_wide <- spread(filtered_long, Indicator.Code, Value)
```

```{r}
# confirming transformation resulted in correct dimensions
length(filter_vector)
cat("\n")
dim(filtered_long)
cat("\n")
dim(filtered_wide)
```

```{r}
head(filtered_wide)
```

## How to Test Data Availability in Terms of Number of Years for a Particlar Indicator.Code and Country

```{r}
# to see year availability need to have column of interest in select() and filter()
filtered_wide %>% select(Country.Name, Year, SL.TLF.TOTL.FE.ZS) %>% filter(Country.Name == 'France', SL.TLF.TOTL.FE.ZS)
```

# Writing to CSV

```{r}
# write.csv(filtered_wide, file = "YOUR_FILENAME.csv", row.names=FALSE)
```




